Further Evidence for Modified Spin-down in Sun-like Stars:\\Pile-ups in the Temperature--Period Distribution

### 2021-12-22 src/ms.tex
paper edits

As discussed in \S\ref{subsec:models} and \S\ref{subsec:rossby}, the existence of a pile-up of stars with a constant Rossby number is consistent with expectations from the WMB model of \citet{vanSaders2016, vanSaders2019}. The location of the long-period pile-up coincides closely with the cluster of main-sequence stars with asteroseismic rotation rates determined by \citet{Hall2021} and we suggest the two features are one and the same. Offsets in both observational and theoretical \teff scales between different studies can explain the modest discrepancies between the asteroseismic sample and the samples considered here, as well as discrepancies in the \rocrit values inferred here and those proposed in \citet{vanSaders2019}. {\color{red} Furthermore, we note that there are important differences between the samples considered here and the \citet{Hall2021} asteroseismic sample. The samples considered here have rotation periods detected from photometric variations and, for a survey of finite baseline and sensitivity like Kepler, photometric rotation periods are harder to detect for slower rotators, smaller amplitude variations, and more stochastic variability patterns. Exacerbating the problem of detection, slower rotation tends to be accompanied by smaller-scale and less periodic variability. The asteroseismic sample, by comparison, is less biased and more likely to contain pile-up stars that our sample may have missed. It is evident from Figure~\ref{fig:gap} that the long-period pile-up also coincides with a steep gradient in variability amplitude, such that a hidden population of pile-up stars may lie just beyond the edge of detectability. This effect, in addition to the \teff offets mentioned above, can further bias our inference of \rocrit to lower values.}

@@

It it also possible that the long-period pile-up studied here only applies to stars with masses $\gtrsim$~1~\msun, in which case the Sun's period and age require no explanation. As seen in Figure~\ref{fig:asteroseismic}, it is not clear that a pile-up exists for \teff~$\lesssim$~5800~K, particularly when examining the CKS and \citet{Hall2021} samples (see also \S\ref{subsec:extent}). However, kernel density estimation of the joint \teff--\prot distribution in the \lamostmcq sample suggests the pile-up may exist at \teff as low as 5500~K or even 5000~K. Further work with a larger sample size may reveal if the pile-up does indeed extend to temperatures as cool as the Sun's.

@@

\begin{enumerate}
    \item We observe an overdensity at the long-period edge of the \teff-\prot distribution of Kepler main-sequence stars with \teff~$\gtrsim$~5800~K. We hypothesize that this pile-up was previously obfuscated by imprecise \teff estimates. Both the existence of the pile-up and its obscuration by large \teff errors were predicted by \citet{vanSaders2019} as a consequence of weakened magnetic braking for stars with $M\gtrsim1$~\msun.

@@

    \item The existence of the long-period pile-up limits the utility of gyrochronology for the hottest stars with convective envelopes, as stellar spin-down appears to stall on the pile-up. For example, a Sun-like star may spend several Gyr evolving across the ridge. Authors using gyrochronology as a means of age-dating a field dwarf star with \teff~$\gtrsim5800~K$ (and possibly cooler \teff as well) should take care to assess whether that star resides on the long-period edge, in which case the uncertainty on the age may be larger than current gyrochronology calibrations imply.

>>>>

As discussed in \S\ref{subsec:models} and \S\ref{subsec:rossby}, the existence of a pile-up of stars with a constant Rossby number is consistent with expectations from the WMB model of \citet{vanSaders2016, vanSaders2019}. The location of the long-period pile-up coincides closely with the cluster of main-sequence stars with asteroseismic rotation rates determined by \citet{Hall2021} and we suggest the two features are one and the same. Offsets in both observational and theoretical \teff scales between different studies can explain the modest discrepancies between the asteroseismic sample and the samples considered here, as well as discrepancies in the \rocrit values inferred here and those proposed in \citet{vanSaders2019}. {\color{red} Furthermore, we note that there are important differences between the samples considered here and the \citet{Hall2021} asteroseismic sample. The samples considered here have rotation periods detected from photometric variations and, for a survey of finite baseline and sensitivity like Kepler, photometric rotation periods are harder to detect for slower rotators, smaller amplitude variations, and more stochastic variability patterns. Exacerbating the problem of detection, slower rotation tends to be accompanied by smaller-scale and less periodic variability. The asteroseismic sample, by comparison, is less biased and more likely to contain pile-up stars that our sample may have missed. It is evident from Figure~\ref{fig:gap} that the long-period pile-up also coincides with a steep gradient in variability amplitude, such that a hidden population of pile-up stars may lie just beyond the edge of detectability. This effect, in addition to the \teff offets mentioned above, can further bias our inference of \rocrit to lower values. Regardless of the biases mentioned above and the true value of \rocrit, the sample considered clearly indicates that the long-period pile-up is apparent for stars with \teff~$>\gtrsim$~\teffmin with Rossby numbers lower than those associated with a detection edge.}

@@

It it also possible that the long-period pile-up studied here only applies to stars with masses $\gtrsim$~1~\msun, in which case the Sun's period and age require no explanation. As seen in Figure~\ref{fig:asteroseismic}, it is not clear that a pile-up exists for \teff~$\lesssim$~5800~K, particularly when examining the CKS and \citet{Hall2021} samples (see also \S\ref{subsec:extent}). However, kernel density estimation of the joint \teff--\prot distribution in the \lamostmcq sample, as well as a cross-match between LAMOST and \citet{Santos2021}, suggests the pile-up may exist at \teff as low as 5500~K or even 5000~K. {\color{red} Furthermore...} Further work with a larger sample size may reveal if the pile-up does indeed extend to temperatures as cool as the Sun's.

@@

\begin{enumerate}
    \item We observe an overdensity at the long-period edge of the \teff-\prot distribution of Kepler main-sequence stars with \teff~$\gtrsim$~5800~K. We hypothesize that this pile-up was previously obfuscated by imprecise \teff estimates. Both the existence of the pile-up and its obscuration by large \teff errors were predicted by \citet{vanSaders2019} as a consequence of weakened magnetic braking for stars with $M\gtrsim$1~\msun.

@@

    \item The existence of the long-period pile-up limits the utility of gyrochronology for the hottest stars with convective envelopes, as stellar spin-down appears to stall on the pile-up. For example, a Sun-like star may spend several Gyr evolving across the ridge. Authors using gyrochronology as a means of age-dating a field dwarf star with \teff~$\gtrsim 5800~K$ (and possibly cooler \teff as well) should take care to assess whether that star resides on the long-period edge, in which case the uncertainty on the age may be larger than current gyrochronology calibrations imply.
